{\bf Practical Concurrency Testing} \\ \normalsize \vspace{1em} \begin{tabular}{c} {\em or: How I Learned to Stop Worrying and Love the Exponential Explosion} \end{tabular}

### 2018-12-24 thesis/quicksand.tex
more revs to quicksand; tweak some figures, phrasings, etc

\begin{itemize}
	\item {\bf Single-state-space model checking} refers to the state-of-the-art model checking strategy,
		i.e., approaching each test with preemption points fixed in advance.
	\item {\bf Maximal state space} refers to the set of thread interleavings possible
		by preempting on all currently-known preemption points,
		whether synchronization or data-race,
		i.e., the singular state space tested by {\em single-state-space model checking}.
	\item {\bf Minimal state space} indicates the opposite:
		those thread interleavings requiring no more than Landslide's mandatory preemptions on voluntary context switches.
	\item {\bf Data-race bug}
		will refer to a concrete failure, such as \Cref{fig:pps-example}'s assertion failure above, whereas
	\item {\bf Data race} refers to the racing access pair itself (\cref{sec:landslide-datarace}).
	\item {\bf Data race candidate} shall refer specifically to
		potentially-racing accesses identified by Limited Happens-Before, %(\cref{sec:landslide-lhb}), % sub sub sec
		when disambiguation with {\em data races} is necessary.
	\item {\bf Benign data races} are those that, when reordered in an alternate interleaving,
		do not lead to a {\em data-race bug}, while
	\item {\bf False positives} are {\em data race candidates} that,
		upon trying to reorder them,
		turn out not to exist in the alternate interleaving at all,
		such as in \Cref{fig:hb-example}(b).
	\item {\bf Nondeterministic data race}
		will refer to data races that cannot be exposed on the first thread interleaving,
		but require model checking to expose to begin with.
\end{itemize}
}

@@

Given unlimited CPU time for testing,
one would always want to switch to the new maximal state space whenever adding a new preemption point.
The maximal state space is guaranteed to subsume all execution sequences reachable in any subset state space,
so considering any incomplete subset of the known preemption points would be redundant work.
\Cref{alg:algorithm0} demonstrates this na\"ive approach.
It is seeded with the set of all statically-known synchronization API preemption points,
and invoked whenever a new data race candidate is found.
%
The upcoming proofs in \cref{sec:quicksand-soundness},
being concerned with the verification guarantee provided when the search may complete within the CPU budget,
are based on this simple version of Iterative Deepening.
The user may also wish to configure her testing tool to prefer this approach, at her discretion,
such as when she believes all bugs have been fixed and wants a verification as fast as possible;
\cref{sec:quicksand-impl-modes} discusses this execution mode further.

@@

For testing user-space code, Quicksand begins with the four
% TODO: decide
jobs from \Cref{fig:id}:
%possible combinations of preemption points from \Cref{fig:pps-statespace}:
$\{yield\}$,
$\{yield,lock\}$,
$\{yield,unlock\}$,
and $\{yield,lock,unlock\}$,
By extension, these also induce preemptions on any other primitives
which use
internal locks,
such as condition variables or semaphores.
Preempting on voluntary switches such as {\tt yield} is always necessary to maintain
Landslide's invariant that only one thread runs between consecutive preemption points,
so the {\tt yield} preemption point is always implicitly enabled.

@@

With Iterative Deepening,
this is a simple matter of creating a new state space
with an additional preemption point enabled on the racing instructions by each thread,
as shown in \Cref{alg:handledatarace}.
I will term these {\em data-race preemption points},
and they will form the foundation of Quicksand's contribution.
Note that even though a data race may involve two different instructions,
$\alpha$ and $\beta$, Quicksand's strategy is to add new state spaces with only one new racing instruction at a time.
Rather than adding a single large state space,
configured to preempt on both involved instructions,
i.e., $AB =$ \GetPPSet{$j_0$} $\cup$ $\alpha$ $\cup$ $\beta$,
it prefers to add multiple smaller jobs which have a higher chance of completing in time, i.e.,
$A =$ \GetPPSet{$j_0$} $\cup$ $\alpha$ and
$B =$ \GetPPSet{$j_0$} $\cup$ $\beta$.
If $A$ and $B$ are bug-free, they will in turn add $AB$ later during their own execution.
The condition on line~1 ensures that we avoid duplicating any state spaces with multiple data-race preemption points;
for example, $AB$ is reachable by multiple paths through its different subsets $A$ and $B$,
but should of course be tested only once.

@@

% TODO: fix syntax hilight
% TODO check figure placement
% TODO check figure camption
\begin{figure}[t]
	\begin{center}
	\begin{tabular}{rll}
		& \multicolumn{2}{c}{\texttt{struct x \{ int foo; int baz; \} *x;}} \\
		& \multicolumn{2}{c}{\texttt{struct y \{ int bar; \} *y;~~~~~~~~~~}} \\
		\\
		& {\bf Thread 1} & {\bf Thread 2} \\
		1 & \texttt{\hilight{brickred}{x->foo = ...;}} & \\
		2 & \texttt{\hilight{olivegreen}{free}(x);} \\
		3 & & \texttt{\hilight{commentblue}{// x's memory reallocated}} \\
		4 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
		5 & & \texttt{\hilight{commentblue}{// ...initialize...}}\\
		6 & & \texttt{publish(y);} \\
		7 & & \texttt{\hilight{brickred}{y->bar = ...;}} \\
	\end{tabular}
	\end{center}
	% TODO: short ver
	\caption{A common execution pattern with {\tt malloc()} that produces false positive data race candidates.}
	\label{fig:recycle}
\end{figure}

@@

% TODO smae as above
\begin{figure}[t]
	\begin{center}
	\begin{tabular}{rll}
		& {\bf Thread 1} & {\bf Thread 2} \\
		1 & \texttt{publish(x);} & \\
		2 & \texttt{\hilight{brickred}{x->foo = ...;}} & \\
		3 & \texttt{\hilight{olivegreen}{free}(x);} \\
		4 & & \texttt{x2 = get\_published\_x();} \\
		5 & & \texttt{\hilight{commentblue}{// x's memory recycled}} \\
		6 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
		7 & & \texttt{\hilight{brickred}{x2->foo = ...;}} \\
	\end{tabular}
	\end{center}
	% TODO: shortver
	\caption{If a single-pass Limited HB analysis discarded candidates matching the malloc-recycle pattern,
it would miss the bug in this adversarial program.}
	\label{fig:recycle-bug}
\end{figure}

>>>>

\begin{itemize}
	\item {\bf Single-state-space model checking} refers to the state-of-the-art model checking strategy,
		i.e., approaching each test with preemption points fixed in advance.
	\item {\bf Maximal state space} refers to the set of thread interleavings possible
		by preempting on all currently-known preemption points,
		whether synchronization or data-race,
		i.e., the singular state space tested by {\em single-state-space model checking}.
	\item {\bf Minimal state space} indicates the opposite:
		those thread interleavings requiring no more than Landslide's mandatory preemptions on voluntary context switches.
	\item {\bf Data-race bug}
		will refer to a concrete failure, such as \Cref{fig:pps-example}'s assertion failure above, whereas
	\item {\bf Data race} refers to the racing access pair itself (\cref{sec:landslide-datarace}).
	\item {\bf Data race candidate} shall refer specifically to
		potentially-racing accesses identified by Limited Happens-Before, %(\cref{sec:landslide-lhb}), % sub sub sec
		when disambiguation with {\em data races} is necessary.
	\item {\bf Data-race preemption point}
		denotes a custom model checker configuration,
		issued after finding a {\em data race},
		requesting it to preempt each involved thread just before its racing memory access
		(\cref{sec:quicksand-dr-pps}).
	\item {\bf Benign data races} are those that, when reordered in an alternate interleaving,
		do not lead to a {\em data-race bug}, while
	\item {\bf False positives} are {\em data race candidates} that,
		upon trying to reorder them,
		turn out not to exist in the alternate interleaving at all,
		such as in \Cref{fig:hb-example}(b).
	\item {\bf Nondeterministic data race}
		will refer to data races that cannot be exposed on the first thread interleaving,
		but require model checking to expose to begin with.
\end{itemize}
}

@@

Given unlimited CPU time, %for testing,
%one would always want to
switching to the new maximal state space whenever adding a new preemption point
\qrevisionminor{would be the quickest way to reach full verification}.
The maximal state space is guaranteed to subsume all execution sequences reachable in any subset state space,
so considering any incomplete subset of the known preemption points would be redundant work.
\Cref{alg:algorithm0} demonstrates this na\"ive approach.
It is seeded with the set of all statically-known synchronization API preemption points,
and invoked whenever a new data race candidate is found.
%
The upcoming proofs in \cref{sec:quicksand-soundness},
being concerned with the verification guarantee provided when the search may complete within the CPU budget,
are based on this simple version of Iterative Deepening.
The user may also wish to configure her testing tool to prefer this approach, at her discretion,
such as when she believes all bugs have been fixed and wants a verification as fast as possible;
\cref{sec:quicksand-impl-modes} discusses this execution mode further.

@@

For testing user-space code, Quicksand begins with the four
%jobs from \Cref{fig:id}:
\qrevisionminor{possible combinations of preemption points from \Cref{fig:pps-statespace}:}
$\{yield\}$,
$\{yield,lock\}$,
$\{yield,unlock\}$,
and $\{yield,lock,$ $unlock\}$, % FIXME argh
By extension, these also induce preemptions on any other primitives
which use
internal locks,
such as condition variables or semaphores.
Preempting on voluntary switches such as {\tt yield} is always necessary to maintain
Landslide's invariant that only one thread runs between consecutive preemption points,
so the {\tt yield} preemption point is always implicitly enabled.

@@

With Iterative Deepening,
this is a simple matter of creating a new state space
with an additional preemption point enabled on the racing instructions by each thread,
as shown in \Cref{alg:handledatarace}.
%I will term these {\em data-race preemption points},
%and they will
These {\em data-race preemption points} form the foundation of Quicksand's contribution.
Note that even though a data race may involve two different instructions,
$\alpha$ and $\beta$, Quicksand's strategy is to add new state spaces with only one new racing instruction at a time.
Rather than adding a single large state space,
configured to preempt on both involved instructions,
i.e., $AB =$ \GetPPSet{$j_0$} $\cup$ $\alpha$ $\cup$ $\beta$,
it prefers to add multiple smaller jobs which have a higher chance of completing in time, i.e.,
$A =$ \GetPPSet{$j_0$} $\cup$ $\alpha$ and
$B =$ \GetPPSet{$j_0$} $\cup$ $\beta$.
If $A$ and $B$ are bug-free, they will in turn add $AB$ later during their own execution.
The condition on line~1 ensures that we avoid duplicating any state spaces with multiple data-race preemption points;
for example, $AB$ is reachable by multiple paths through its different subsets $A$ and $B$,
but should of course be tested only once.

@@

% TODO: fix syntax hilight
% TODO check figure placement
% TODO check figure camption
\begin{figure}[h]
	\begin{center}
	\begin{tabular}{rll}
		& \multicolumn{2}{c}{\texttt{struct x \{ int foo; int baz; \} *x;}} \\
		& \multicolumn{2}{c}{\texttt{struct y \{ int bar; \} *y;~~~~~~~~~~}} \\
		\\
		& {\bf Thread 1} & {\bf Thread 2} \\
		1 & \texttt{\hilight{brickred}{x->foo = ...;}} & \\
		2 & \texttt{\hilight{olivegreen}{free}(x);} \\
		3 & & \texttt{\hilight{commentblue}{// x's memory reallocated}} \\
		4 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
		5 & & \texttt{\hilight{commentblue}{// ...initialize...}}\\
		6 & & \texttt{publish(y);} \\
		7 & & \texttt{\hilight{brickred}{y->bar = ...;}} \\
	\end{tabular}
	\end{center}
	\caption[Reallocation false positive pattern.]
	{Reallocation false positive pattern arising from
	{\tt malloc()} returning the same memory that was just {\tt free()}d.}
	\label{fig:recycle}
\end{figure}

@@

% TODO smae as above
\begin{figure}[h]
	\begin{center}
	\begin{tabular}{rll}
		& {\bf Thread 1} & {\bf Thread 2} \\
		1 & \texttt{publish(x);} & \\
		2 & \texttt{\hilight{brickred}{x->foo = ...;}} & \\
		3 & \texttt{\hilight{olivegreen}{free}(x);} \\
		4 & & \texttt{x2 = get\_published\_x();} \\
		5 & & \texttt{\hilight{commentblue}{// x's memory recycled}} \\
		6 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
		7 & & \texttt{\hilight{brickred}{x2->foo = ...;}} \\
	\end{tabular}
	\end{center}
	\caption[Reallocation-pattern data race that hides a true bug.]
	{Reallocation-pattern data race that hides a true bug.
	If a single-pass Limited Happens-Before analysis discarded all data race candidates
	with intervening reallocations,
	it would miss the bug in this adversarial program.}
	\label{fig:recycle-bug}
\end{figure}
